### 2019-07-11 manuscript.tex
progress on Laurentia tectonic history section

Laurentia refers to the craton that forms the Precambrian core of North America (Fig. \ref{fig:Laurentia_map}). Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne+Nain provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collison recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior province with the last evidence of that orogeny being ca. 1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper \textit{United Plates of America} \citep{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the more major cratonic blocks: the Thelon orogen associated with the collision between the Slave province and the Rae province ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; and the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and Nain provinces \citep{St-Onge2009a}. Crustal growth also progressed at this time within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal and undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks arrived ca. 1.72 Ga \citep{Kilian2016a} which is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgmation of Archean provinces with the Trans-Hudson orogen further to the northeast.

@@

@@

Accretionary orogenesis continued along the south to southeast margin of Laurentia with Elzevirian

The ca. 1190 to 1160 Ma Shawinigan orogeny is interpreted to be due to the accretion of a
terrane comprised of amalgamated arc volcanics and associated metasediments \citep{McLelland2010a}

In the latest Mesoproterozoic ca. 1.11-1.08 Ga, a major intracontinental rift co-located with the a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior craton and Paleoproterozoic provinces. This Midcontinent Rfit lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.  Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) which is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a} and is typically to interpreted to have been a major collisional event associated with the amalgamation of continents into the supercontinent Rodinia \citep{Hoffman1991a}.

This tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from recorded from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.06 Ga) requires an long-lived open active margin up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}. This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but is sometimes other continental blocks are reconstructure into positions (e.g. Amazonia in \citealp{Elming2009a}) that are seemingly incompatible with this record of accretionary orogenesis. The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself requires a collision between Laurentia and (an)other continent(s) ca. 1080 Ma -- the geological observation of which first lead to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.

There was significantly less crustal growth on the western margin of Laurentia and the tectonic history is not as well elucidated as on the southern to eastern margin. A thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca. 1.47 to 1.40 rifting \citep{}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

@@

@@

Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean cratons. There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain cratons. Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.

For the Superior craton, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there as substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}. \cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga. I follow this interpretation and group the poles into Superior (West) and Superior (East). These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame.

Following Laurentia's amalgamation, poles from each part of the Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Hall201Xa} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of thick plateau developed during continent-continent collision (e.g. \cite{}). Poles with a B-rating are also included in the composite that come from terranes that were once part of Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction. for which I used the following rotations

@@

>>>>

Laurentia refers to the craton that forms the Precambrian core of North America (Fig. \ref{fig:Laurentia_map}). Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne+Nain provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collison recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collision orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave province and the Rae province ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and Nain provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to
1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the Nain province \citep{St-Onge2009a}. As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior with still goingd undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks arrived ca. 1.72 Ga \citep{Kilian2016a} which is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgmation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga lead to rapid amalgamation of the majority of the Laurentia craton.

@@

Crustal growth also progressed at this time through accretionary orogenesis. This accretion occured within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior province with the last evidence of that orogeny being ca. 1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.

@@

Accretionary orogenesis continued along the (south)east margin of Laurentia with the arc-continent collision of the ca. 1.25–1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a
terrane comprised of amalgamated arc volcanics and associated metasediments \citep{McLelland2010a}. In the latest Mesoproterozoic ca. 1.11-1.08 Ga, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior craton and Paleoproterozoic provinces. This Midcontinent Rift lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.  Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) which is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.

There was significantly less crustal growth on the western margin of Laurentia and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin. A thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca. 1.47 to 1.40 rifting which is typically interpretted to be intracratonic\citep{}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.06 Ga) requires an long-lived open margin without a major conjugate continent up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}. This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but is sometimes other continental blocks are reconstructured into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself requires a collision between Laurentia and (an)other continent(s) ca. 1080 Ma -- the geological observation of which first lead to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. That the Laurentia margin experienced large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny that is recorded in Texas, up through the Blue Ridge Appalachian inliers, through Ontario and up to the Labrador Sea remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting. Along the western margin of Laurentia small-scale rifting occured ca. 780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountins of NW Laurentia \citep{Rooney2017a}. However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until near the ca. 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a}. The emplacement of the Gunbarrel large igneous province along this margin and the extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin initiating ca. 780 Ma. If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the late Ediacaran. The geological evidence therefore supports active tectonism along the western margin of Laurentia, but suggests that more dramatic lithospheric thinning occured later than the timing of rifting typically implemented in models of Rodinia break-up. It is conceivable that the rifting that drove latest Ediacaran to Cambrian thermal subsidence was associated with a microcontinent fragment rather than a major conjugate continent, but it is an important constraint for paleogeographic models that the time of major lithospheric thinning on the western Laurentia margin is ca. 539 Ma. The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin. Available geochronological constraints and thermal subsidence modeling indicate ca. 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a,Halverson2018a}. These data suggest that conjugate continental lithosphere rifted away from northeast Greenland at the time. Extensive rifting that was followed by thermal subsidence occured along the southeast to east Laurentia margin at the Neoproterozoic-Phanerozoic boundary. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen) as well as long term Cambrian to Ordovician subsidence along the margin. The Argentine pre-Cordillera Cuyania terrane is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting. As with other rifts, it is difficult to distinguish the seperation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar. One interpretation is that there was break-up along the eastern margin that initationed ca. XXX associated with the Rome trough and prior to the ca. 539 Oklahoma aulacogen rifting that liberated Cuyania.

@@

Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic provides very strong evidence for mobile plate tectonics driving Laurentia evolution throughout the past 2 billion years. This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic rcord.

@@

Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean cratons. There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces. These data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly. Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.

For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}. This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny. \cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga.  I follow this interpretation and group the poles into Superior (West) and Superior (East). These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame. Uncertainty remains with respect to whether the ca. 1.88 Ga Molson dikes pole pre-dates or post-dates this rotation and thus for the time-being should be considered solely in the western Superior province reference grame.

Following Laurentia's amalgamation, poles from each part of the Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Hall201Xa} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of thick plateau developed during continent-continent collision (e.g. \cite{}). Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland. These terranes were once part of contigous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the following rotations:

@@

To make an argument that differential plate tectonic motion began in the Neoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
